---
title: "The dot symbol"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library("purrr")
library("magrittr")
library("data.table")
```

The dot `(.)` appears in different places in the R ecosystem: purrr, magittr pipe. I will explore and explain what happens if you mix these usages, or nest them, how the dot symbol is special and how it is not.

## Basic usage

### %>% of magrittr

You have to use the dot if the parameter you pipe forward is not the first parameter of your next function or if you use pipe with data.table and use `[]` function.

```{r}
lipsum::lipsums[1] %>% 
    stringi::stri_trans_totitle() %>% 
    gsub(' ', '-', .)
```

You can also refer to the parameter with `.` if it is the first, but then you do not have to:

```{r}
lipsum::lipsums[1] %>%
    stringi::stri_trans_totitle(.) %>% 
    stringr::str_replace_all(., ' ', '-')
```

### Map of purrr

```{r}
calendar_consts <- list(
    'num_hours_in_day' = 24,
    'num_days_in_week' = 7,
    'num_months_in_year' = 12
)
map_chr(names(calendar_consts), ~ stringr::str_c(., ': ', calendar_consts[[.]]))
```

For compact anonymous functions the formula notation with `.` is a shorthand for the more verbose following known from the base R `apply` family.

```{r}
map_chr(names(calendar_consts), function(name) {stringr::str_c(name, ': ', calendar_consts[[name]])})
```

Of course defining your function outside the call to map is always possible and preferable for more complex functions.

## Nested usage

### Pipe within pipe

I could not came up with a realistic usage for this as pipe is exactly for avoiding nesting...

### Map within map

```{r}
consts <- list(
    'calendar' = calendar_consts,
    'geo' = list(
        'num_continents' = 7,
        'num_states_in_US' = 50
    )
)
map(
    consts,
    ~{
        sub_list <- .
        map_chr(
        names(sub_list),
        ~ stringr::str_c(., ': ', sub_list[[.]])
    )}
)
```

`.` is actually not that different from ordinary variable names which means that scoping rules apply as usual. In the inner if you refer to `.` it means the current element in the inner map. You can either save the outer current element to a variable or define the desired environment for your variable name.

```{r}
map(
    consts,
    ~ map_chr(
        names(.),
        ~ stringr::str_c(., ': ', parent.env(environment())$'.'[[.]])
    )
)
```

While the above is possible I do not recommend it as it is difficult to read.
You could instead do the following:

```{r}
pasteConstNamesAndValues <- function(const_list) {
    map_chr(
        names(const_list),
        ~ stringr::str_c(., ': ', const_list[[.]])
    )  
}
map(consts, pasteConstNamesAndValues)
```


## Mixed usage

### Pipe in map

### Map in pipe

