---
title: "the dot symbol"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library("purrr")
library("magrittr")
library("data.table")
```

The dot `(.)` appears in different places in the R ecosystem: purrr, magittr pipe. I will explore and explain what happens if you mix these usages, or nest them, how the dot symbol is special and how it is not.

## Basic usage

### %>% of magrittr

You have to use the dot if the parameter you pipe forward is not the first parameter of your next function or if you use pipe with data.table and use `[]` function.

```{r}
lipsum::lipsums[1] %>% 
    stringi::stri_trans_totitle() %>% 
    gsub(' ', '-', .)
```

You can also refer to the parameter with `.` if it is the first, but then you do not have to:

```{r}
lipsum::lipsums[1] %>%
    stringi::stri_trans_totitle(.) %>% 
    stringr::str_replace_all(., ' ', '-')
```

### map of purrr



## Nested usage

## Mixed usage


